### 2024-08-16 paper.tex
dimension
delete a bunch of stuff that I don't need anymore!

\par Now we are only left with finding the matrices \(A \in C_2\) satisfying \(Ab = b\).
This just means finding the matrices \(A\), of the form given above, which satisfy \(Ab = 0\).
We may restate this as the condition that for every \(i\),
\[\begin{pmatrix}
    A_{i1} & \cdots & A_{in}
\end{pmatrix}b = 0.\]
Taking the transpose, 
\[b^\top \begin{pmatrix}
    A_{i1}^\top\\
    \vdots\\
    A_{in}^\top
\end{pmatrix} = 0.\]
Now, the rightmost column of \(A_{ij}^\top\) is \(v_{ij}\) upside-down (rows reversed).
Write \(u_{ij}\) to refer to \(v_{ij}\) upside-down, and \(u_{ij}'\) to refer to \(v_{ij}'\) upside-down.
Then the second-rightmost column is \(u_{ij} \upshift 1\), and so on until the first column is \(u_{ij} \upshift [\lambda_i - 1]\).
So, we can reformulate this condition by requiring that for all \(l\),
\[b^\top \begin{pmatrix}
    u_{i1} \upshift l\\
    \vdots\\
    u_{in} \upshift l
\end{pmatrix} = 0.\]
Writing
\[b = \begin{pmatrix}
    b_1\\
    \vdots\\
    b_k
clearly this is equivalent to requiring that for all \(l\),
\[\begin{pmatrix}
    b_1^\top \rightshift l & \cdots & b_k^\top \rightshift l
\end{pmatrix} \begin{pmatrix}
    u_{i1}\\
    \vdots\\
    u_{ik}
\end{pmatrix} = 0.\]
Then writing \(b_j = \begin{pmatrix}
    b_j'\\
    \vdots
\end{pmatrix}\), where \(b_j'\) has the same dimension as \(u_{ij}\), we can rewrite this constraint as 
\[\begin{pmatrix}
    b_1'^\top \rightshift l & \cdots & b_k'^\top \rightshift l
\end{pmatrix} \begin{pmatrix}
    u_{i1}'\\
    \vdots\\
    u_{ik}'
\end{pmatrix} = 0.\]
Finally, requiring the above equation for all \(l\) is the same thing as requiring
\[\begin{pmatrix}
    b_1'^\top \rightshift 0 & \cdots & b_k'^\top \rightshift 0\\
    \vdots &  & \vdots \\
    b_1'^\top \rightshift [\lambda_i - 1] & \cdots & b_k'^\top \rightshift [\lambda_i - 1]
\end{pmatrix} \begin{pmatrix}
    u_{i1}'\\
    \vdots\\
    u_{ik}'
\end{pmatrix} = 0.\]
This is the most that we will simplify the constraint.
Recall that the \(v_{ij}'\) can be chosen arbitrarily, so the \(u_{ij}'\) can be chosen arbitrarily as well.
That is, the set \(C_3 := \{A \in \gl_m : AX = XA, aA = a, Ab = b\}\) is the set of \(I + A\) for any \(A\) that can be recovered from \(u_{ij}'\) meeting the above condition.
\par Now we find the dimension of \(C_3\).
Since the \(u_{ij}'\) can be chosen arbitrarily, apart from the condition above, we just need to add up their dimensions and subtract the rank of the matrix constructed of the \(b\)'s.
The dimension of \(u'_{ij}\) is \(\min(\lambda_i, \lambda_j)\) in the case that \(i \neq i_0\), and \(\min(\lambda_i - j_0, \lambda_j)\) when \(i = i_0\).
Write \(d_{ij}\) for this dimension.
For each \(i\), let \(q_i\) be the length of the longest prefix of \(b_i\) which consists of zeroes.
Then if \(A_{X, a, b}\) is nilpotent, we must have \(q_{i_0} \geq j_0\).
So we may as well set \(q_{i_0} := q_{i_0} - j_0\).
\par Fix an \(i\).
I claim the rank of the matrix of \(b\)'s is \(h := \max_j [d_{ij} - q_j]\), or zero if this quantity is always negative.
First, to see that the rank is at least this big, note that for any \(j \in \argmax_j [d_{ij} - q_j]\), the rank of the \(j\)th ``column block'' of the matrix of \(b\)'s is equal to \(h\).
Then, to see that the rank is at most this big, note that every row after the first \(h\) rows is just zeroes.
So, the total degrees of freedom for the \(u_{ij}\) is 
\[\sum_j d_{ij} - \max(0, \max_j [d_{ij} - q_j]).\]
Then summing over all the \(i\), we get the dimension of \(C_3\):
\[\sum_i \left[\sum_j d_{ij} -\max(0, \max_j [d_{ij} - q_j])\right] = \sum_{ij} d_{ij} - \sum_i \max(0, \max_j[d_{ij} - q_j]).\]
When \(i \neq i_0\), then \(d_{ii} = \lambda_i \geq q_i\), and therefore \(\max(0, \max_j [d_{ij} - q_j]) = \max_j [d_{ij} - q_j]\).
When \(i = i_0\), then \(d_{ii} = \lambda_i - j_0 \geq q_{i_0}\).
So, we can simplify the above expression:
\[\sum_{ij} d_{ij} - \sum_i \max_j [d_{ij} - q_j].\]
Note that \(\max_j [d_{ij} - q_j] = \max_j [\min(p_i + q_i, p_j + q_j) - q_j]\).
This thing being maximized equals \(\min(p_i + q_i - q_j, p_j)\).
It would be really helpful if I could simplify this to \(p_j\) or something.
I don't think so, though.

\section{A necessary condition for \(A_{X, a, b}\) to be nilpotent}
Suppose \(A := A_{X, a, b}\) is nilpotent.
Define the \emph{height} of a vector \(v \in \mathbb{C}^{m + n}\) as the smallest \(k \geq 0\) such that \(A_{X, a, b}^k v = 0\).
Clearly, for \(i \leq j\), we have that the height of \(f_i\) is geq the height of \(f_j\).
Thus, for each \(k \geq 0\), we must have \(A_{X, a, b}^k f_n \in \langle e_{ij} \rangle_{ij}\).
We have \(Af_n = \sum_{ij} b_{ij} e_{ij}\).
Then \(A^2 f_n = \sum_{ij} b_{ij} (e_{i, j - 1} + a_{ij} f_1)\), so we must have \(\sum_{ij} b_{ij} a_{ij} = 0\).
Similarly, since \(A^3 f_n \in \langle e_{ij}\rangle\), we see that \(\sum_{ij} b_{ij} a_{i, j - 1} = 0\).
Continuing in this way, we obtain that for all \(k \geq 0\),
\begin{equation}\label{ab_nilp}\sum_{ij} b_{ij} a_{i,j - k} = 0.\end{equation}
In fact this is in fact a sufficient condition for \(A_{X, a, b}\) to be nilpotent, as can be seen from the characteristic polynomial of \(A_{X, a, b}\).
However, we will not do this calculation here, instead just showing that it is sufficient by giving a Jordan basis.

\section{A Jordan basis for \(A_{X, a, b}\)}
For illustration, we will begin with the case that \(X\) is a single Jordan block.
Then we will move to the general case.

\subsection{When \(X\) is a Jordan block}
Suppose \(X\) is a single Jordan block of size \(m\).
In this case we just write the basis of \(\mathbb{C}^m\) as \(e_1, ..., e_m\), and the above condition on \(a\) and \(b\) simply becomes
\begin{equation}
    \forall k \geq 0. \; \sum_{i = k + 1}^m b_i a_{i - k} = 0.
\end{equation}
This condition can be simplified even more.
\begin{lemma}
    The condition (1) holds iff there exist nonnegative \(m_1, m_2, m_3\), \(a_1, ..., a_{m_3}\), and \(b_1, ..., b_{m_3}\) satisfying the following conditions.
    \begin{itemize}
        \item \(m = m_1 + m_2 + m_3\)
        \item \(a = (0, 0, ..., 0, a_1, ..., a_{m_3})\)
        \item \(b = (b_1, ..., b_{m_1}, 0, 0, ..., 0)\)
        \item If \(m_1 \neq 0\), then \(a_{m_1} \neq 0\)
        \item If \(m_3 \neq 0\), then \(b_1 \neq 0\)
    \end{itemize}
\end{lemma}
\begin{proof}
    If \(a\) and \(b\) are of this form, then for every \(k \geq 0\) we have \(b_i a_{i - k} = 0\), so clearly the condition holds.
    \par Now suppose (1) holds.
    If \(a\) or \(b\) is zero, this is trivial.
    Otherwise, let \(m_1 = \max\{i : a_i \neq 0\}\), and let \(m_3 = \min\{i : b_i \neq 0\}\).
    We just need to show that \(m_1 < m_3\).
    For contradiction, suppose \(m_1 - m_3\) is nonnegative.
    Then by (1), \(0 = \sum_i b_i a_{i - (m_1 - m_3)} = b_1 a_{m_1}\), contradicting that \(b_1\) and \(a_{m_1}\) are nonzero.
\end{proof}

@@

\subsection{The general case}
In the case where \(X\) was a single Jordan block, it was helpful to have the condition that all the nonzero \(a\)'s were to the right of all the nonzero \(b\)'s.
This allowed some tricks with right-shifting.
In the general case, the parameters \(a\) and \(b\) are not necessarily in such a helpful form.
We will see, however, that we can change basis of \(\mathbb{C}^m\) to put them in a nice form.
\par Recall our nilpotent operator \(A_{X, a, b}\) on \(\mathbb{C}^m \times \mathbb{C}^n\).
We can define a map \(\mathbb{C}^m \to \mathbb{C}\) by \(x \mapsto x \cdot a\).
Then, by the previous lemma, we can choose a Jordan basis \(e_{ij}\) of \(\mathbb{C}^m\) so that at most one \((i, j)\) has \(f(e_{ij}) \neq 0\).
This implies that there are some \(P, k, b'\) such that \(PA_{X, a, b}P^{-1} = A_{X, 1_k, b'}\).
So, without loss of generality, we may consider this case, where \(a = 1_k\).
\par Write our basis of \(\mathbb{C}^m\) as \(e_{ij}\) and our basis of \(\mathbb{C}^n\) as \(f_j\).
Let \(i_0, r\) be such that \(a = e_{i_0, \lambda_{i_0} - r + 1}\).
Let \(b' = A^{n + r} e_{i_0, \lambda_{i_0}} = b + e_{i_0, \lambda_{i_0} - r - n}\).
Let \(b_{ij}'\) be the projection of \(b'\) onto \(e_{ij}\).
For each \(i\), let \(p_i = \max\{j : b_{ij}' \neq 0\}\).
Then for each \(i \neq i_0\), let \(q_i = \lambda_i - p_i\).  Let \(q_{i_0} = \lambda_{i_0} - r - p_{i_0}\).

@@

\par We have two cases, which will closely resemble the cases we had when \(X\) was a single Jordan block.
\subsubsection{First case: \(n + \max_i p_i \geq q_{i_0} + p_{i_0}\)}
We list the chains in (roughly) decreasing order.

@@

TODO: do this.

\subsubsection{Second case: \(n + \max_i p_i \leq q_{i_0} + p_{i_0}\)}
Again we list the chains in (roughly) decreasing order.
Note that, since \(n + \max_i p_i \leq q_{i_0} + p_{i_0}\), in particular we have \(n \leq q_{i_0}\), and \(n \leq q_i\) for all \(i \leq i_0\).
Now, for all \(i \neq i_0\) with \(q_i \geq n\), take the chain of length \(p_i + q_i\) starting with \(e_{i,p_i + q_i}\).
Then, take the chain starting with \(e_{i_0, \lambda_{i_0}}\), of length \(\lambda_{i_0}\).
\par Let \(i_1 = \min\{i : n > q_i\}\).
Let \(P = \max_{i \geq i_1} p_i\).
Then take the chain beginning with \(f_1 - (A^{n + p} f_1 >> n + p)\), of length  \(n + P\).
\par Finally, we handle \(i \geq i_1\).  
(In this case \(q_i < n\).)
If \(p_i \leq \max_{k > i} p_k\), then take the chain beginning with \(e_{i, p_i + q_i}\) of length \(p_i + q_i\).
Otherwise, set 
\[v_i = A^{n - q_i} f_1 - \sum_{k = 1}^{p_i} b_{i,k}' e_{i, q_i + k}.\]
Note that \(A^{q_i} v_i\) is just \(b'\) with row \(i\) zeroed out.
Now, we take the chain beginning with \(v_i - (A^{q_i + \max_{k > i} p_k} v_i >> q_i + \max_{k > i} p_k)\), of length \(q_i + \max_{k > i} p_k\).
\par TODO: where did I use the fact that \(n + \max_i p_i \leq q_{i_0} + p_{i_0}\).

>>>>

\section{A Jordan basis for \(A_{X, a, b}\)}

@@

\subsection{The general case}
Let \(X\) be nilpotent in Jordan form, and let \(a = \mathbbm{1}_{w,1}\), and let \(b\) be such that \(\forall j. \; b_{wj} = 0\).
We describe a Jordan basis for \(A_{X, a, b}\).

\par Let \(\mu\) be the result of removing the row \(w\) from \(\lambda\).
For each row \(i\), let \(p_i = \max\{j : b_{ij} \neq 0\}\) (the maximum of the empty set is zero).
Then set \(q_i = \lambda_i - p_i\).  Let \(q_{i_0} = \lambda_{i_0} - r - p_{i_0}\).

@@

We list the chains in (roughly) order of decreasing length.

@@

Indeed, the sum of the lengths is
\[\left[\sum_{i : q_i \geq n + r} (p_i + q_i)\right] + (r + n + \max_{i : q_i < n + r} p_i) + \sum_{i : q_i < n + r} \begin{cases}
    p_i + q_i, & p_i \leq \max_{k > i} p_k \\
    q_i + \max_{k > i} p_i, & p_i > \max_{k > i} p_k
\end{cases} = \]
\[\left[\sum_{i : q_i \geq n + r} (p_i + q_i)\right] + (r + n + \max_{i : q_i < n + r} p_i) + \sum_{i : q_i < n + r} (q_i + \min(p_i, \max_{k > i} p_i)).\]
Let \(i_1 < \cdots < i_s\) be the `peaks': that is, the set \(I_1 = \{i : q_i < n + r \land p_i > \max_{k > i} p_k\}\).
(As I should've mentioned before, we take the max of the empty set to be zero.)
Let \(I_2 = \{i : q_i < n + r\} \setminus I_1\).
Then 
\[\sum_{i : q_i < n + r} (q_i + \min(p_i, \max_{k > i} p_i)) = \sum_{i \in I_2} (p_i + q_i) + \sum_{k = 1}^s (q_{i_k} + p_{i_{k + 1}}).\]
Clearly then, the whole sum is \(m + n\).
